---
title: "Competition II Day 3"
author: "Roma Siugzdaite"
date: "2025-05-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Competition II. Factor analysis

This dataset is about Student Habits vs Academic Performance

This is a simulated dataset exploring how lifestyle habits affect academic performance in students. With 1,000 synthetic student records and 15+ features including study hours, sleep patterns, social media usage, diet quality, mental health, and final exam scores, it’s perfect for ML projects, regression analysis, clustering, and data viz. Created using realistic patterns for educational practice.

Ever wondered how much Netflix, sleep, or TikTok scrolling affects your grades? 👀 This dataset simulates 1,000 students' daily habits—from study time to mental health—and compares them to final exam scores. It's like spying on your GPA through the lens of lifestyle. 

student_habits_performance.csv

#Research question: How many groups I can see in these data? 


```{r, echo=TRUE }

library(readr)
pairfam <- read_csv("C:\\Users\\cgranadamunoz\\OneDrive - Universiteit Antwerpen\\Codes_UA\\multivariate_statistics\\pairfam_synth_data.csv")
head(pairfam)

```

## Descriptive statistics

```{r, echo=TRUE }

# select only numerical variables
data <- pairfam[, c(3, 5:13)]
summary(data)

data <- as.data.frame(data)

#Check missing values
#Where is this Na's are? use this function
colSums(is.na(data))
missings <- colSums(is.na(data))
summary(missings)

```

## Normality testing

```{r, echo=TRUE }

# I create a function to make a figure of a certain size
fig <- function(width,heigth){
    options(repr.plot.width = width, repr.plot.height = heigth)
}

# show all histograms in one figure

fig(25,25)
par(mfrow=c(2,5)) # 2x4 grid for variable visualization
for (i in seq(1,ncol(data))){
    hist(data[,i],main = '',xlab = colnames(data)[i])
}


```

What correlations tell you??? 
There two main factors 

```{r, echo=TRUE}

library(car)
fig(25,25)

par(mfrow = c(2,5))
for (i in seq(1,ncol(data))){
    qqPlot(as.numeric(data[,i]),ylab=colnames(data)[i])
}
```

## Shapiro test for normality

Shapiro Test
H0 = Normal Distribution
H1 = Not normal distribution

We can create a table of all results and keep the test values.

```{r}
normal_uni_test <- data.frame(matrix(0,ncol(data),3))
rownames(normal_uni_test) <- colnames(data)
colnames(normal_uni_test) <- c('Statistic','p-value','Normal')

alpha <- 0.05
for (i in seq(1,ncol(data))){
    normal_uni_test[i,1] <- shapiro.test(as.numeric(data[,i]))$statistic
    normal_uni_test[i,2] <- shapiro.test(as.numeric(data[,i]))$p.value
    if (normal_uni_test[i,2] < alpha){
        normal_uni_test[i,3] <- 'NO'
    } else{
        normal_uni_test[i,3] <- 'YES'
    }
}


#write.csv(normal_uni_test, file = 'univariate_normal_test.csv')
View(normal_uni_test)
```

## Correlation


```{r}

# calculate and plot correlations
library(RcmdrMisc)
library(psych)
r <- rcorr.adjust(data, type = "pearson", use = "complete.obs")
corPlot(r$R$r, numbers = TRUE, stars = TRUE, diag = FALSE)

#other method
library(ggplot2)
library(corrplot)

fig(20,20)
r <- cor(data, method = "spearman")
corrplot(r, type = 'upper', method = 'square')


# png(filename = 'corrplot.png',bg = 'transparent',width = 2000,height = 2000,pointsize = 30)
# corrplot(cor(df),type = 'upper',method = 'ellipse')
# dev.off()

```
## Clean data
```{r}
library(RcmdrMisc)

data2 <- na.omit(data)

r <- rcorr.adjust(data2, type = "pearson", use = "complete.obs")
corPlot(r$R$r, numbers = TRUE, stars = TRUE, diag = FALSE)

# you can use also a new library, example here
library(ggplot2)
library(corrplot)
r <- cor(data2, method = "spearman")
corrplot(r, method = 'square')

```
## Bartlet test

```{r}
summary(data2)
library(psych)
# Bartlet's test of Sphericity
cortest.bartlett(data2)
bart.test <- cortest.bartlett(r, n = nrow(data2))
bart.test$p.value # extract p-value
```

Bartlet test has to be significant < 0.05 

## Scree plot

```{r}
fa.parallel(data2,fm='fa',ylabel='Eigenvalues')

```
## Factor analysis

The way we learned in the class


```{r}

Nfacs <- 4 # This is for four factors. You can change this as needed.

library(stats)
fit <- factanal(data2, Nfacs, rotation="varimax", correlation= TRUE) #We need to put correlation TRUE if we want to use correlation. 

print(fit, digits=2, cutoff=0.3, sort=TRUE)
```

Reporting: There are 2 factors explaining 30% of total variance. With first factor explaining 20% and second factor 10% of variance. Factor one eigenvalue is 1.79 and factor two eigenvalue is 0.89.


```{r}

library(psych)
KMO(data2)

```

KMO has to be more than 0.6 for factor analysis. We have 0.17. :( Indicating data are not clustered inside the dataset.


```{r}

factor_analysis <- fa(data,nfactors=4,fm='ml',rotate='varimax',max.iter=100)
factor_analysis

```


```{r}

# Name your latent factors. if you have two factors - your have t give to names. If you have 3 factors, you have to give 3 names, and ect.
colnames(factor_analysis$loadings) <- c('m_health','Internet_use', 'Satsfaction','Depresion')
fig(20,20)
fa.diagram(factor_analysis,rsize = 0.5,cex=0.01)
```

Your interpretation about the data? 

```{r}
#data2 <- data[,c(2,3, 7:9)]
factor_analysis <- fa(data2,nfactors=4,fm='ml',rotate='varimax',max.iter=100)
factor_analysis
```

```{r}
colnames(factor_analysis$loadings) <- c('Studying', 'Environment')
fig(20,20)
fa.diagram(factor_analysis,rsize = 0.5,cex=0.01)
```

## Report to th paper results

## New example

```{r}
library(readr)
data <- read_csv("/Users/rsiugzdaite/Downloads/pairfam_synth_data.csv")
head(data)
```
```{r}
Data <- data[,c(3, 5:13)]
Data <- data.frame(Data)
par(mfrow=c(2,5)) # 2x4 grid for variable visualization
for (i in seq(1,ncol(Data))){
    hist(Data[,i],main = '',xlab = colnames(Data)[i])
}

```


```{r}

library(ggplot2)
library(corrplot)

r <- cor(Data, use = 'pairwise.complete.obs')
corrplot(r, method = 'square')

```


```{r}
fa.parallel(Data,fm='fa',ylabel='Eigenvalues')
```

```{r}

DataClean <- na.omit(Data)


library(psych)
KMO(DataClean)
DataClean2 <- DataClean[, KMO(DataClean)$MSAi > 0.49] # Get rid of all variables with MSA < 0.50

Nfacs <- 4 # This is for four factors. You can change this as needed.

library(stats)
fit <- factanal(DataClean2, Nfacs, rotation="varimax")

print(fit, digits=2, cutoff=0.3, sort=TRUE)

KMO(DataClean2)
```


```{r}
# number of loadings is the number of factors
load <- fit$loadings[,1:Nfacs]
load
plot(load,type="p") # set up plot
text(load,labels=names(DataClean2),cex=.7)
```


```{r}
library(psych)

loads <- fit$loadings
loads
fa.diagram(loads)

```
## Your interpretation? 
