---
title: "MultivariateAnalysis"
output:
  html_document: default
  pdf_document: default
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Beer data


cost is representing..


```{r}

```
We are importing the 'beer.txt' data. Please check and change your path accordingly. The default folder is the github repository "C:\Users\cgranadamunoz\OneDrive - Universiteit Antwerpen\Codes_UA\multivariate_statistics"


```{r }
#Import the data. Change to your own path!!
Beer <- read.csv("C:\\Users\\cgranadamunoz\\OneDrive - Universiteit Antwerpen\\Codes_UA\\multivariate_statistics\\Beer.txt")
#Visualization of the data. 2 ways...
View(Beer)
head(Beer, 5) #By default head is showing only 6 rows of the file

#Descriptive statistics. First check which type of data we have.
summary(Beer)
#Where is this Na's are? use this function
colSums(is.na(Beer))
missings <- colSums(is.na(Beer))
summary(missings)

#visualization of the data to determine normality. 3 methods
hist(Beer$cost)
#if in shapiro p value is very low, is not normal
shapiro.test(Beer$cost)
#look the dispersion of the data. Association...
plot(Beer$cost, Beer$reputat)

```

## Including Plots

You can also embed plots, for example:
```{r}

```

```{r, echo=FALSE}

library(RcmdrMisc)

rcorr.adjust(Beer)

Beer2 <- na.omit(Beer)
r <- rcorr.adjust
corPlot(r$R$r,numbers = TRUE, stars = TRUE, diag = FALSE, val = r$R$P)
rcorr.adjust(Beer2)

```

## KMO Test
Kaiser provided the following values for interpreting the results:

* 0.00 to 0.49 unacceptable
* 0.50 to 0.59 miserable
* 0.60 to 0.69 mediocre
* 0.70 to 0.79 middling
* 0.80 to 0.89 meritorious
* 0.90 to 1.00 marvelous

```{r}

library(psych)
KMO(Beer2)

```
```{r}

BeerClean <- Beer2[, KMO(Beer2)$MSAi > 0.515] # Get rid of all variables with MSA < 0.50

KMO(BeerClean)$MSAi
round( KMO(BeerClean)$MSA, 2 )

```

```{r}
library(psych)
cortest.bartlett(BeerClean)
#If p-value is less than p<0.05 we reject null hypothesis there are no common factors in my data. We accept alternative hypothesis, that are common factors. 
```
PCA to establish the numbers of PCA by the elbow rule


```{r}
scree(Beer, pc=FALSE) 
scree(BeerClean, pc=FALSE) 

```
```{r}
ev <- eigen(cor(BeerClean)) # get eigenvalues
ev$values
```

```{r}
fa.parallel(BeerClean, fa="fa")
```




```{r}
Nfacs <- 3 # This is for factors. You can change this as needed.

fit <- factanal(Beer2, Nfacs, rotation="promax")
#You can change the rotation, the default is Varimax

print(fit, digits=2, cutoff=0.3, sort=TRUE)
#Uniquenesses need to be low values
```

```{r}
load <- fit$loadings[,1:2]
plot(load,type="p") # set up plot
text(load,labels=names(BeerClean),cex=.7)

```
```{r}
library(psych)

loads <- fit$loadings

fa.diagram(loads)
```


```{r}

dim(fit$loadings)

round(fit$loadings[1:6,], 2)
```


```{r}
FactorLoadings <- round(fit$loadings[1:6,], 2)

write.csv(FactorLoadings, file="FacLoads.csv")
```


