---
title: "Day3Script"
author: "Roma Siugzdaite"
date: "2024-03-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## ###################################
## Correspondence analysis (CA)
## ###################################

Read the data 

```{r, echo=TRUE }

europeans <- read.csv("europeans.txt", stringsAsFactors=TRUE)

```

## Contingency table

```{r , echo=TRUE}

library("FactoMineR")
library("factoextra")

europeans.table <- xtabs(freq ~ country + character, data=europeans)
mosaicplot(europeans.table, color = (1:6), main = "character vs image")

```

```{r}

library(ISLR)
library(tidyverse)
library(Rfast)
library(MASS)


europeans.tbl.stand <- prop.table(europeans.table)
addmargins(europeans.tbl.stand)

```
Statistical Tests

We can apply the following statistical tests in order to test if the relationship of these two variables is independent or not.

Chi-Square Test

We have explained the Chi-Square Test in a previous post. Let’s run it in R:
```{r}

chisq.test(europeans.tbl.stand)


```
As we can see the p-value is more than 5% thus we can accept the null hypothesis that the image is dependent to brand.!!!!

Log Likelihood Ratio

Another test that we can apply is the Log Likelihood Ratio using the MASS package:



```{r}
loglm( ~ 1 + 2, data = europeans.tbl.stand) 
```
Again, we accepted the null hypothesis.

```{r}

CA(europeans.tbl.stand, ncp = 5, graph = TRUE)

res.ca <- CA(europeans.tbl.stand, graph = FALSE)

eig.val <- get_eigenvalue(res.ca)
eig.val

fviz_screeplot(res.ca, addlabels = TRUE, ylim = c(0, 65))

```



```{r}
# repel= TRUE to avoid text overlapping (slow if many points)
fviz_ca_biplot(res.ca, repel = TRUE)
```

Use the function fviz_ca_row() [in factoextra] to visualize only row points:

```{r}

fviz_ca_row(res.ca, repel = TRUE)

```

```{r}
# Color by cos2 values: quality on the factor map
fviz_ca_row(res.ca, col.row = "cos2",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), 
             repel = TRUE)
```

```{r}
library("corrplot")

row <- get_ca_row(res.ca)
row
corrplot(row$cos2, is.corr=FALSE)


corrplot(row$cos2, is.cor = FALSE, tl.cex = .7, method = "color")

```

```{r}
# Cos2 of rows on Dim.1 and Dim.2
fviz_cos2(res.ca, choice = "row", axes = 1:2)
```
It’s also possible to create a bar plot of rows cos2 using the function fviz_cos2().

```{r}
col <- get_ca_col(res.ca)
col

fviz_ca_col(res.ca, col.col = "cos2", 
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE)
```
```{r}
fviz_contrib(res.ca, choice = "col", axes = 1:2)
```

The R code below draws a standard asymmetric biplot:

```{r}
fviz_ca_biplot(res.ca, 
               map ="rowprincipal", arrow = c(TRUE, TRUE),
               repel = TRUE)
```


```{r}
fviz_ca_biplot(res.ca, map ="colgreen", arrow = c(TRUE, FALSE),
               repel = TRUE)
```
